Data Specialization for a Staged Lambda Calculus

### 2015-02-24 paper/intro.tex
small edits all over
working on the intro

@@

Previous work on data specialization has been limited to simple, imperative
languages. In this paper, we extend data specialization to a typed lambda
calculus, allowing us to specialize more interesting programs.

>>>>

Consider the following function that takes as arguments an $x$ and $m$
$y$'s and invokes a function $f(\cdot, \cdot)$ on consecutive
values of $y$'s.
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
  @$(f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@.
\end{lstlisting}
Let's assume for concreteness that $f(x,y)$ takes time to the size of
the first argument.  The function $F(x_0, y_0 \ldots y_{m-1}$ then
takes time $O(n \cdot m)$ where $n$ is the size of the input $x_0$.

Such computations, where we apply a multivariate function to arguments
where only one of the arguments change over time, are common.  In
fact, many research problems revolve around improving performance of
such functions.  For example, algorithm design techniques such as
dynamic programming take advantage of the fact that the function $f$
above is being passed the same argument many times, and thus optimize
it so that it can re-use results related to $f$.  Similarly, 
crucial compiler optimization techniques such as loop hoisting and
common subexpression evaluation take identify common common
computations across all invocations of $f$ and perform them once and
for all at the beginning of the computation.

The common theme in all such techniques is to achieve what Jorring and
Scherlis back in 1980's called: {\em frequency reduction} and {\em
  precomputation}~\cite{JS86-staging}.  The idea behind frequency
reduction is to identify computations that are performed multiple
times and pull them ahead so that they can be performed once and used
later as needed.  The idea behind precomputation is to identify
computations that can be performed earlier, for example at compile
time if their inputs are available statically and perform them at that
earlier time.  Jorring and Scherlis discuss three possible approaches
to achieving frequency reduction and procomputation by performing {\em
  stage transformations} that stages the computation into distinct
stages where the earlier stages host computations that can be
performed earlier or that can might be used frequently.  The three
approaches are
\begin{itemize}
\item 
partial evaluation,

\item
meta programming, and

\item 
pass separation.

\end{itemize}

The idea behind partial evaluation is to take advantage of the fact
that certain data becomes available earlier than others.
Specifically, partial evaluation distinguishes between a {\em static}
and {\em dynamic} stage and performs all static computations first,
followed be the dynamic stage.  For example, suppose that we wish to
compute $F(x_0, y_0 \ldots y_{m-1})$ and that $x_0$ is a static
variable becoming available at an earlier stage.  We can take
advantage of this by specializing $f$ for $x_0$ such that $f(x,y) =
f_{x0}(y)$, where $f_{x0}$ is the function specialized for $x0$.  We
can then re-write our function $F$ as follows.

\begin{lstlisting}
fun @$F_{x0}(y_0, y_1 \ldots y_{m-1}$@ = 
  @$(f_{x0}(y_0);  f_{x0}(y_1)  \ldots f_{x0}(ym)$@.
\end{lstlisting}

There has been much research on partial evaluation (\secref{related}).
Broadly speaking, earlier (pre 90's) research used binding-time
analysis to determine the static and dynamic parts of a computation
and specialized the program based on that.  Binding time analysis can
be fully automated but can be difficult to predict.  Later research
tends to use {\em meta programming}, where the programmer writes a
program that given the static values generates a specialized program
for those values.  Meta programming gives the programmer relatively
fine-grained control over partial evaluation but comes at the cost of
more complex reasoning and a loss of automation.

The third and final approach, called pass separation, assumes much
less information to perform a stage transformations.  Instead of
requiring the static data, it requires just the knowledge of which
computations are static and dynamic and uses this knowledge to
generate a stage one (static) and stage two (dynamic) computations.
Applied to our example, pass separation would {\em split} function
into $f1$ and $f2$ such that $f(x,y) = f2(f1(x),y)$.  We can then
rewrite our example as
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
@$(f2(f1(x), y_0);  f2(f1(x), y_1)  \ldots f2(f1(x), y_{m-1})$@.
\end{lstlisting}
%
A compiler can then rewrite this as 
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
let @$fx = f1(x)$@
in @$(f2(fx, y_0);  f2(fx, y_1)  \ldots f2(fx, y_{m-1})$@.
\end{lstlisting}
We refer to such a program as {\em untangled}.

%
The key difference with partial evaluation is that we have not
specialized the program to any specific static value but simply
translated it into an equivalent program where the first stage
function $f1$ only consumes first-stage data.  We can evaluate such an
expression in two stages: first, we evaluate all $f1(x)$'s and reduce
them to a final value that can be used directly in the second stage
(we can be smart about this evaluation and evaluate $f1(x)$ just
once). Second, we evaluate all instances of $f2$'s and compute the
final value.

Since pass separation requires strictly less information than partial
evaluation and meta programming, it can be a powerful technique that
can be applied with far less information.  For example, the code that
we obtained in our example, can be instantiated with a value of $x$
that becomes available at run time and still obtain significant
benefits.  As we show in this paper, these benefits can be
asymptotically significant.  In other words, we can obtain some
benefit of partial evaluation at run time in cases when we don't know
the values of arguments ahead of time.  If, however, we do know the
values ahead of time, then we can also apply partial evaluation to the
untangled program, deriving again all the benefits of partial
evaluation.

As the example illustrates, untangling the program into two stages
seems to have no cost.  So why don't we apply this kind of untangling
all time time?

\smallskip

Untangling a program into stages to can be done manually.  In fact
Jorring and Sherlis describe how to do this for some examples.
Automation, however, has been difficult.  According to Jorring and
Sherlis ``... partial evaluation technique is readily mechanized,
while staging transformation approach will elude full automation for
some time.''

\smallskip

That time has come.

In this paper...

\paragraph{Old intro below.}

@@

Previous work on data specialization has been limited to simple, imperative
languages. In this paper, we extend data specialization to a typed lambda
calculus, allowing us to specialize a broader class of programs.
